---
title: "Naive Bayes"
output: html_document
---

#Naive Bayes Example
##Load Data

```{r}
train <- read.csv("in_class_nb.csv", header=TRUE)
summary(train)
```

We don't have much to clean up in our data so we can jump to the 
model phase.

#Construct Model

```{r}
library("e1071")
#This will reproduce our spreadsheet calculations
bm <- naiveBayes(sport ~ height + endurance, data=train)
bm
#This is how we want to run it when we're working with our data
#bm <- naiveBayes(sport ~ height + endurance, data=train, laplace=1)
#bm
```

##Evaluate Model

```{r}
train$class <- predict(bm, newdata=train, type="class")
train$preds <- predict(bm, newdata=train, type="raw")
```

When we evaluate our models, we run our predictions against the 
training data so we can compare our predictions to the known classes.

##Evaluate the Results

```{r}
results <- table(train$class,train$sport)
results
bb_total_pred <- sum(results[1,])
bb_total_class <- sum(results[,1])
bb_precision <- results[1,1]/bb_total_pred
bb_recall <- results[1,1]/bb_total_class

soc_total_pred <- sum(results[2,])
soc_total_class <- sum(results[,2])
soc_precision <- results[2,2]/soc_total_pred
soc_recall <- results[2,2]/soc_total_class

accuracy <- (results[1,1]+results[2,2])/sum(results)
```

Confusion Matrix `r sum(results)`   |Condition Basketball|Condition Soccer|Total_Predicted|Precision
-------------------|--------------------|------------------|---------------|---------
__Predicted Basketball__|`r results[1,1]`|`r results[1,2]` |`r bb_total_pred`|`r bb_precision`
__Predicted Soccer__|`r results[2,1]`    |`r results[2,2]` |`r soc_total_pred`|`r soc_precision`
__Total_Condition__| `r bb_total_class`  | `r soc_total_class`|               |
__Recall__         |`r bb_recall`        |`r soc_recall`       |              |

The accuracy is `r accuracy`

##Deployment

Let's make our predictions on new data.

```{r}
test <- read.csv("in_class_nb_test.csv", header=TRUE)
test$class <- predict(bm, newdata=test, type="class")
test$preds <- predict(bm, newdata=test, type="raw")
test
```