---
title: "Explanatory Modeling with Linear Regression and Factors"
output: html_document
---

#Explaining Low Birth Weight with Linear Regression
##Understanding the Problem
You work at a small hospital with a limited amount of equipment for treating 
new borns with low birth weight. The hospital administrator has asked you to help 
use data to understand low birth weight in your hospital's population. If you find a compelling explanation, you might decide to adjust your model to predict birth weights based on a number of characteristics. 

Since you know about linear regression, you have the tools to give this a shot. You are provided with the following data.

##Understanding the Data

The following data set is available from:
https://www.umass.edu/statdata/statdata/data/

Hosmer and Lemeshow. (2000). Applied Logistic Regression. Second Edition. John Wiley and Sons. 

The metadata description of the fields is available here:
https://www.umass.edu/statdata/statdata/data/clslowbwt.txt

```{r}
data <- read.csv("low_birth_weight_train.csv", header=TRUE)
str(data)
```

The training data contains the following fields:

Field       |Notes
------------|----------------------------------------
ID          |Patient ID to be excluded from analysis
LOW         |Summary variable to be excluded 
AGE         |Age of the expecting mother in years
LWT         |Weight of the expecting mother in pounds
RACE        |1=white; 2=black; 3=other
SMOKE       |0=non-smoker; 1=smoker
PTL         |To be excluded
HT          |To be excluded
FTV         |To be excluded
BWT         |Birthweight of newborn in grams

##Data Preparation
###Remove fields to be excluded from analysis

The following code removes all fields except AGE, LWT, RACE, SMOKE, and BWT.

```{r}
ndata <- data[,c(3,4,5,6,11)]
str(ndata)
```

###Factors

We have two variables that are not numbers in any meaningful sense: RACE and SMOKE.
In order to handle these properly, we need to convert them to factors.
Factors represent categorical data in an appropriate format for linear regression and 
other forms of analysis.
At this stage, it's important to recognize why RACE and SMOKE must be coded 
as factors. Currently, RACE includes three values: 1,2,3 representing 
patients described as WHITE, BLACK or OTHER respectively. Let's forget (for now) the woefully inadequate model of race and ethnicity represented in this encoding. We cannot 
leave these values as is because any measures of centrality we use will tend toward 
2 which represents BLACK. There is no reason to encode 2 with this 
category: this is nominal data after all. Therein lies the problem: if we shuffle the categories encoding OTHER with 
the value of 2, then we will change the meaning of our average. In other words, whatever we code as 2 will end up being the center of the "distribution". These simply aren't numbers.

Luckily, we can fix this rather easily.

```{r}
ndata$RACE <- factor(ndata$RACE)
ndata$SMOKE <- factor(ndata$SMOKE)
str(ndata)
```

We can now see that R knows to treat these as categorical variables instead of 
integers. We'll see why this is so important a bit later. 

##Modeling

We want to predict birth weight (BWT) based on our other variables.
So we might start with a model that looks something like this.

$BWT = m_1 * AGE + m_2 * LWT + m_3 * RACE + m_4 * SMOKE + b$

We create this formula in R with the following code:

```{r}
fm <- lm(BWT ~ AGE + LWT + RACE + SMOKE, data=ndata)
fm 
```

The linear model we created suggests the following formula:

$BWT = `r round(fm$coefficients[2],3)` * AGE + `r round(fm$coefficients[3],3)` * LWT + `r round(fm$coefficients[4],3)` * RACE2
+ `r round(fm$coefficients[5],3)` * RACE3 + `r round(fm$coefficients[6],3)` * SMOKE1 + `r round(fm$coefficients[1],3)`$

Let's walk through this model step by step.

* Every year in a mother's age tends to decrease a newborn's birth weight by 2 grams.
* Every pound of a mother's weight tends to increase a newborn's weight by 4 grams.
* If an expecting mother is described as BLACK (RACE=2), then this tends on average to decrease a newborn's weight by 511 grams.
* If a woman is described as OTHER (RACE=3), then this tends on average to decrease a 
newborn's weight by 400 grams.
* If a woman smokes, then this tends on average to decrease a newborn's weight by 
400 grams.

Now for a prediction:
We have a 21 year old expecting mother who weighs 100 pounds and is described as white 
and a non-smoker. What prediction would our model give for the weight of her newborn?

We would fill in our formula as follows:

$`r fm$coefficients[2] * 21 + fm$coefficients[3] * 100 +  fm$coefficients[4] * 0 + fm$coefficients[5] * 0 + fm$coefficients[6] * 0 + fm$coefficients[1]` = 
`r round(fm$coefficients[2],3)` * 21 + `r round(fm$coefficients[3],3)` * 100 + `r round(fm$coefficients[4],3)` * 0
+ `r round(fm$coefficients[5],3)` * 0 + `r round(fm$coefficients[6],3)` * 0 + `r round(fm$coefficients[1],3)`$

```{r}
fm$coefficients[2] * 21 + fm$coefficients[3] * 100 +  fm$coefficients[4] * 0 + fm$coefficients[5] * 0 + fm$coefficients[6] * 0 + fm$coefficients[1]
```

So our model predicts that this woman's child will be about 3200 grams at birth.

##Analysis and Evaluation

So far so good, but how good is our model? Let's take a look.

```{r}
summary(fm)
```

First, let's check the significance of our model coefficients.
We can see that most of our coefficients are significant with p-values 
less than 0.05. The mother's age, however, is __NOT__ statistically significant.

Although we have a statistically significant model fit (p-value = .00001744), the 
model does not fit particularly well (Adjusted R-squared = 0.1251).

As an explanatory model, this is pretty unsatisfying, but this might 
still be a useful model for prediction.

At this point, we might want to see if we can improve our model by removing 
noisy features like the mother's age (AGE). We could do this manually by 
removing some features from the original data set and refitting the model.
This process is called backwards elimination. R provides a handy way of doing 
this automatically.

```{r}
rm <- step(fm, direction = "backward", trace=FALSE ) 
summary(rm)
```

The new model increases the fit slightly by removing AGE.
Although this model ultimately explains a small part of the variation
seen in infant birth weights, we might still use this model for prediction because 
it contains the information available to us at the time we want to 
make our decision.

##Deployment: Making Predictions

Now we want to use our best model to make predictions with new data. In this 
scenario, we will allocate scarce equipment based on our estimates of 
birth weight.

###Load Test Data

```{r}
test <- read.csv("low_birth_weight_test.csv", header=TRUE)
str(test)
```

###Prepare Data

We want to remove the ID field.

```{r}
test <- test[,2:4]
```

We also need to convert RACE and SMOKE into factors.

```{r}
test$RACE <- factor(test$RACE)
test$SMOKE <- factor(test$SMOKE)
str(test)
```

###Calculate our predictions

Let's make some predictions on our new data.

```{r}
test$preds <- predict(rm, newdata=test)
```

The `predict()` takes the newdata (test) and predicts birth weight with the linear model (rm).

```{r}
test
```

###Reviewing the predictions

Let's recreate a few predictions manually and think them through.

In test case 1, we have an expecting mother described as a black, non-smoker weighing 182 pounds. Our model predicted that her newborn will weigh 3011 grams.

$`r rm$coefficients[2] * 182 +  rm$coefficients[3] * 1 + rm$coefficients[4] * 0 + rm$coefficients[5] * 0 + rm$coefficients[1]` = 
`r round(rm$coefficients[2],3)` * 182 + `r round(rm$coefficients[3],3)` * 1
+ `r round(rm$coefficients[4],3)` * 0 + `r round(rm$coefficients[5],3)` * 0 + `r round(rm$coefficients[1],3)`$

Now let's look at test case 6. We have a 75 pound woman described as a black, non-smoker weighing 75 pounds. Our model predicted that her newborn will weigh 2590 grams.

$`r rm$coefficients[2] * 75 +  rm$coefficients[3] * 1 + rm$coefficients[4] * 0 + rm$coefficients[5] * 0 + rm$coefficients[1]` = 
`r round(rm$coefficients[2],3)` * 75 + `r round(rm$coefficients[3],3)` * 1
+ `r round(rm$coefficients[4],3)` * 0 + `r round(rm$coefficients[5],3)` * 0 + `r round(rm$coefficients[1],3)`$

Both predictions are based on the desired calculations, but the prediction for test case 6 has a serious problem. The prediction is based on a mother's weight that is outside of the range used in the training data. In other words, our prediction is extrapolating beyond the range of the independent variables providing the basis of our model. We should recognize when we are doing this and think deeply about whether this is a good idea or not. In many business applications, this is a bad idea because we are making a prediction with no empirical support at that range. In scientific applications, we may sometimes have good reasons for doing this. 

```{r}
summary(test[,c("LWT")])
```

Based on our predictions, we recognize that cases 3 and 4 carry the highest risk of low birth weight. Given what we know now, we would allocate our resources to those cases. While case 6 also represents risk, our model's ability to predict this case is severely limited.

##Reflection

In this simple model, we see how variables such as weight, smoking and race/ethnicity affect birth weight. Since our model explains relatively little of the variance in the dependent variable, we obviously have many things that we can do to improve our process. First, there are probably many other variables that will provide relevant information including other medical conditions, nutrition information and other behavioral indicators. Second, the categories for race/ethnicity may not be sufficiently detailed to capture health information for all populations. Keep in mind that the other category might include people from places as varied as China, India, and Egypt. The other category as currently constructed may be hiding or aggregating risk profiles that are quite different. Moreover, white and black may be insufficient for describing a person's background. Even smoker vs. non-smoker may not be accurate enough. We could just as easily ask a person how many cigarettes he or she smokes in a typical day.   
