---
title: 'Chapter 1: Understanding R'
output: 
    tufte::tufte_html:
      highlight: tango
---

#A Gentle Introduction to R

```{marginfigure}
Chapter Summary:
This chapter introduces the R programming language and 
the foundations required for using it to solve data-oriented problems.
```

Many students find the R programming language troubling. Beginners find the 
syntax baffling and arcane. More experienced programmers also express 
frustration with R because it is different from many commonly-used programming 
languages. Basically, both groups of students struggle with 
R because it asks them to think about their work in new ways. This chapter 
introduces students to the basics of R programming. First, the philosophy 
behind R is explained to help students orient themselves to its assumptions and nuances.
Second, the chapter explains the rationale for using R to teach data-oriented
topics to undergraduates. Finally, the chapter discusses RStudio, an 
integrated development environment for R, R programming basics, and some
common errors students encounter when working with R.

##The R Philosophy

The R philosophy includes a few assumptions about programming that 
may be unfamiliar even to experienced students of technology. The R
philosophy can be summed up with a few simple phrases.

* Data analysis is the primary purpose of R.
* Programming is a means to an end in R.
* Flexibility is more important than consistency in R.

Statisticians and mathematicians developed R to meet their specific 
disciplinary needs. Initially, the primary audience for the software 
consisted of highly trained academics and their students with specific
analytical needs. For this reason, R is intensely focused on tasks related 
to data analysis. Many other programming languages like Java, for example, 
focus on the development of applications and programs. If a student's 
experience is primarily with Java, R will feel strange because it wasn't 
designed for creating applets, command line programs or interactive 
graphical applications. R was made for data analysis, and therefore, will 
lack many of the features familiar to those with experience in languages 
for application development.

Many programming languages like Java were created by people who really like 
writing programs and developing applications. Moreover, such programming 
languages were written to be as general as possible. In theory, a person 
can write a program in Java to address pretty much any kind of application.
Languages like Java have been used for games, web sites, business applications,
and anything else one might imagine. In contrast to Java, R was 
written to serve people who thought of programming as a means to an end. 
Statisticians and mathematicians are less interested in programming for its 
own sake than professional programmers. They focus instead on what programs can 
do for them such as help them to complete their analytical tasks more 
quickly and accurately.

Since the R community has been developing its packages for over 20 years, 
the language has grown a great deal. While many programming languages 
focus on the consistency of the code, R is more concerned with flexibility. In other words, R developers try to ensure that 
its users have as much freedom as possible. This is somewhat 
different from the philosophy of other programming languages. For example, 
Java frequently retires old versions of its packages requiring older 
programs to be updated to use its new libraries. Because R sacrifices 
consistency for flexibility, R can often feel 
unnecessarily complicated.

For the purposes of this course, students should understand and embrace the 
R philosophy. This course teaches students about data analysis. The programs
students will write in this course are a means to that end. Although R can 
feel complicated at times, once students get the hang of it, they will be 
able to achieve a great deal with rather compact programs.

##Rationale for using R

If R is such an unusual language, then why do we use it for this course?
Despite its challenges, R has some important advantages that make its use 
compelling for this course.

* R is free.
* R works on almost any desktop computer with Windows, Mac or Linux.
* R is powerful and awesome once people buy into its philosophy.
* R has been adopted by many disciplines in the sciences, social sciences, 
  humanities and arts.
* R developers have contributed packages addressing an impressive array of 
  real-world applications.
  
Basically, R provides the best balance available for the modern student. 
It is used 
widely in academic and professional settings. It is used across the great 
breadth of disciplines pursued by undergraduate students today. In short, 
R provides the best platform to address the diverse needs of 
the undergraduate students who typically enroll in this course.

##Mathematics and Statistics

Given the early history of R, it is no surprise that R has powerful features 
for handling mathematics and statistics. For example, suppose a student 
wants to explore the correlation between two variables: temperature in 
degrees Celsius and ice cream sales in US dollars[^1].

[^1]: Pierce, Rod. (5 Jan 2016). "Correlation". Math Is Fun. 
Retrieved 10 Apr 2016 from http://www.mathsisfun.com/data/correlation.html

```{r echo=FALSE}
temp <- 
  c(14.2, 16.4, 11.9, 15.2, 18.5, 22.1, 19.4, 25.1, 23.4, 18.1, 22.6, 17.2)
sales <- c(215,325,185,332,406,522,412,614,544,421,445,408)

data <- data.frame()
data <- data.frame(t(rbind(temp,sales)))
```

Consider the following sample data:

```{r echo=FALSE}
knitr::kable(
  data, caption = 'Ice Cream Sales vs Temperature'
)
```

For this example, we want to know if people buy more ice cream when the 
weather is warmer. Statisticians use a correlation coefficient to measure 
the relationship between two variables. If the two variables are positively 
correlated, then the correlation coefficient will be close to 1. If the 
two variables are negatively correlated, the correlation coefficient will be
close to -1. Finally, if the variables are uncorrelated, then the 
correlation coefficient will be close to 0. R makes this kind of calculation
fairly straight-forward.

```{r}
cor(data, method="pearson")
```

The sample code above uses the `cor()` function to calculate the 
correlation coefficient. In this example, the table of values was 
stored in a variable called `data`. The `cor()` function calculates 
the correlation coefficient using a technique called Pearson's product 
moment correlation coefficient, `method="pearson"`.

The output of this command is printed directly below the command. The 
correlation coefficient in this case is `r cor(data$temp,data$sales)`.
In other words, based on this sample data we have some evidence supporting
the idea that people buy more ice cream when it's hot outside.

##Graphics

Quite frankly, if this is all R did for us, we would not use it for this class.
Excel has similar functions that work well for this kind of task.
But R makes it possible for us to explore such relationships graphically.
For example, it is common to use a scatter plot
when displaying the relationship between two variables

```{r fig.margin=TRUE, fig.cap="Ice Cream Sales vs. Temperature"}
plot(data$temp,
     data$sales,
     xlab="Temperature in Degrees Celcius",
     ylab="Sales in U.S. Dollars")
```

The code above creates a scatter plot with the temperature data plotted 
on the x axis and sales data plotted on the y axis. The example labels 
the x and y axes appropriately.

##Data Science Libraries

In addition to R's mathematical and graphical abilities, R also provides 
many packages for data science algorithms that we will discuss in this 
class. For example, we will use packages to support the following 
data mining topics in this course.

* Data Mining Packages
    + Decision Trees
    + Neural Networks
    + Support Vector Machines
    + Naive Bayes Classification
* Network Science Packages

It is difficult to find software with the great range of capabilities 
offered by R.

#R Basics

R is free. RStudio is free. You can obtain them from the following websites.

Main R website: https://www.r-project.org/

Download R: http://cran.mirrors.hoobly.com/

Main RStudio website: https://www.rstudio.com/products/rstudio/

Download RStudio: https://www.rstudio.com/products/rstudio/download3/

##RStudio

RStudio is a popular integrated development specifically for R programming. 
Although R can be used without R Studio, most R developers use R Studio 
because it provides many important services to help with the development 
process. For an introduction to R Studio, watch the short video outlining 
some of its useful features.

##R Syntax

R commands are called _functions_. Perhaps the most important function in R
is the `help()` function. The `help()` function provides additional information
about any other function in R.

Functions have _arguments_ which control the specific aspects of how a 
function is executed. For example, if we want to read the help file for 
the `cor()` function, we would call the help function with `cor` as its
argument.

```
help(cor)
```
Arguments always appear between the parentheses in an R function.

In many cases, we will want to use more than one argument in a function.
Arguments are separated by commas in R. For example, if we decide to 
specify that we want to see the help file for the `cor()` function in the 
`stats` package, we would type:

```
help(cor, package="stats")
```
In some cases, arguments take the form of `argument="option"` or 
`argument=option`.

The generic form of R functions, therefore, is:

```
function(argument1, argument2=option, ...)
```

In addition to functions, R allows users to define _variables_.
Variables store values in our programs. For example, if we want to 
store the name "Al" into a variable called `who`, we could type:

```{r}
who <- "Al"
```

The `<-` symbol is referred to as the assignment operator. We use this to 
assign a value to a variable.

We type the name of the variable to see the value it stores.

```{r}
who
```

R has many _operators_ for mathematical purposes including arithmetic and 
boolean operators.

With functions and operators, we will be able to construct complex 
programs for data analysis.

#Common Tasks in R

##Set Working Directory

Since we will be using R to analyze data stored in files, we will need to be
able to perform file management tasks easily. This requires the user to 
understand the concept of a _working directory_. The working directory in R 
is the location in the computer's file system where R is currently looking 
for files. If we want to see the current working director in R, we use 
the `getwd()` function:

```{r}
getwd()
```

Now that we know where R is looking for our files, we might want to know 
what it sees. To print the file names in the current working directory, 
we can use the `dir()` function:

```
dir()
```
In many cases, we will need to change the working directory in R. We use 
the `setwd()` function for this purpose. We can change the working directory 
to the `C:\temp` folder with the following function:

```{r}
setwd("C:/temp")
```

You should become very comfortable with these functions because we will 
frequently need to change directories to load data into R.

##Loading Data

Since R is designed for data analysis, it is no surprise that it has many 
functions for loading data. First, R already comes with many built in 
data sets. To explore the data sets already available in R, use the 
`data()` function.

If you want to load a specific data set such as the `cars` data set, call the
`data()` function with `cars` as its only argument.

```{r}
data(cars)
```

This command loads the `cars` data set and stores it in a variable named
`cars`.
To see the data stored in the varialbe, just type `cars`.

```{r}
cars
```

Although this is extremely convenient, analysts often need to load data from 
files or the internet. R provides many ways to do this. For the sake of 
clarity, we will discuss one particular file type, the comma separated 
value file or CSV for short. The `read.csv()` function loads data in 
CSV format into R. For example, if we want to load data from the following 
location over the internet, http://introcs.cs.princeton.edu/java/data/elements.csv,
we would type:

```{r}
data <- read.csv("http://introcs.cs.princeton.edu/java/data/elements.csv")
```

This function downloads the CSV from internet and stores its contents into 
a variable called `data`.

If we downloaded the same file to the `C:\temp` folder, we would load it 
with a slightly different command:

```
data <- read.csv("C:/temp/elements.csv")
```

Practice loading data into R. We will be performing this task often.

##Installing and Loading Packages

In many cases, developers need to load R _packages_, software modules that 
extend R's capabilities. First, packages must be installed into the R system.
After installation, packages must be loaded into memory to make them 
available to R programs.

To install the `stringr` package, for example, we type:

```
install.packages("stringr")
```

To make `stringr` functions available in R, type:

```
library(stringr)
```

The stringr functions and documentation are now available in R. To test this,
look at the help documents for the function.

```
help(package="stringr")
```

Since many data mining tasks require additional libraries, you should practice
installing and loading libraries to master this skill.

#Common Errors in R

Many students express frustation due to errors in R. In this section, we 
highlight a few common errors to help smooth out the rough patches.

##Error: could not find function "gtest"

```
gtest()
```

This error can mean a few different kinds of things. Literally, R does not 
know of any function called `gtest()`. There are a few different reasons 
why this might be:

1. There is in fact no such function because you mistyped something.
Review the documentation and double-check your typing.

2. This is such a function, but its packages has not been installed and loaded.
In this case, install and load the package and try again.

##Error in file(file, "rt") : cannot open the connection

Students often receive this error when first working with data files.

```
data <- read.csv("bacon.csv")
```

The key is to read the entire message. R tells you that there is no such 
file or directory. In this case, R cannot find the file `bacon.csv`. 
There are several reasons why this may be so.

1. The file in fact does not exist.
Confirm that you correctly typed the file name. Confirm that the file is in
fact where you think it is.

2. The file exists, but is not in R's current working directory.
Move the file to R's working directory or change R's working 
directory to the location where the file can be found.

#Summary

In this chapter, we outlined the basics of working with R including its 
philosophy, basic syntax and common tasks. In the next chapter, we will 
introduce the data mining concepts that form the primary focus of this 
course.

#References

R Core Team (2016). R: A language and environment for statistical
  computing. R Foundation for Statistical Computing, Vienna, Austria. URL
  https://www.R-project.org/.